Potential competition in alpine bumble bees is mitigated by imperfect trait-matching and subpartitioning within floral morphotypes

### 2021-03-10 ms_2/ms_2_v2.Rmd
back to lab computer

@@

All analyses were conducted in `R` `r citep(citation())`. Data handling and visualization was performed with the `tidyverse` suite of R packages [@Wickham2019-aa]. Bayesian multilevel models were implemented in Stan [@Stan2020aa] via the package `brms` [@Burkner2017-aa] using default priors, and visualizations were prepared with the packages `tidybayes` `r citep(citation("tidybayes"))` and `ggdist` `r citep(citation("ggdist"))`. Ordination analysis and permutation testing were performed with the package `vegan` `r citep(citation("vegan"))` and visualized using the packages `ggplot2` [@Wickham2016-aa] and `ggvegan` [@Simpson2019aa]. Hierarchical generalized additive models (HGAMs) were implemented with the package `mgcv` [@Wood_2017] and visualized with the packages `mgcViz` [@Fasiolo2018aa] and `tidymv` [@Coretta-aa2021]. All R code used in this analysis, together with full model output and diagnostics, is available in Appendix X.

@@

<!-- State more explicityl how the models should be interpreted with respect to bias -->

@@

To investigate elevational patterns of *within-site* niche overlap, we modeled community-mean niche overlap as a function of level (morphotype, family, genus) and elevation using HGAMs. Community mean niche overlap was calculated by taking the mean niche overlap (within level) across all species pairs for each site. Level was included as both a fixed intercept effect and a smoothing factor in conjunction with elevation, following the type "I" model form described by @Pedersen2019-kf. To account for the proportional nature of our data, we used a beta distribution family with a logit link function.

@@

```{r fig3, echo = FALSE,  fig.cap = paste("Posterior conditional mean visitation probability by Kugler morphotype and bumble bee tongue length class. Point-interval plots depict 66\\% and 95\\% CIs. Letters indicate group differences based on evidence ratio at alpha = 0.05. Note that the range of the x-axis has been rescaled for each panel to emphasize the comparison of visitation probabilities across tongue length classes.")}
knitr::include_graphics("../output/brm_pvis00_plot1.png")
```

```{r fig4, echo = FALSE,  fig.cap = paste("Posterior conditional mean visitation probability by Kugler morphotype and bumble bee species. The tongue length class of each species is color-coded. Point-interval plots depict 66\\% and 95\\% CIs. Note that the range of the x-axis has been rescaled for each panel to emphasize the comparison of visitation probabilities across bumble bee species")}
knitr::include_graphics("../output/brm_pvis02_plot1.png")
```

@@

While bumble bees are, indeed, remarkably generalized in their foraging, they are highly discriminating generalists, dramatically over-representing certain floral taxa and morphotypes and ignoring others. In our study system, lip, flag, and head flowers dominated overall visitation probability when visitation was pooled across all bumble bee species, while bell, disc, and stalk-disc flowers were underrepresented. Patterns of pooled visitation are, of course, driven disproportionately by the behavior of the most abundant bumble bee species, which in our system were *B. pascuorum*, *B. pratorum*, *B. soroeensis*, *B. terrestris/lucorum*, and *B. wurflenii* (Sponsler et al., submitted). Visitation to lip and flag flowers was driven especially by the medium-tongued *B. pascuorum*, the most abundant bumble bee in our study system, and secondarily by the medium-tongued *B. wurflenii*. Visitation to head flowers was driven mainly by the short-tongued *B. soroeensis*.

@@

<!-- If not tongue length, what else could explain the variation in floral morphotype preference witin tongue length classes? -->

@@

The mystery of bumble bee species coexistence hinges on the relationship between species-level floral preferences, trait-matching, and the emergent phenomenon of community-level patterns of diet overlap. While tongue length is a relatively strong predictor of species-level floral preferences, it is at best a weak predictor of diet overlap. This is due in part to divergent preference of bumble bees with similar tongue lengths toward certain floral morphotypes. Perhaps more importantly, though, diet overlap at the level of floral genus tended to be about half as strong as diet overlap at the level of floral morphotype, indicating that bumble bees with similar morphological affinities, driven in part by tongue length compatibility, subdivide floral morphotypes at the genus- and species-level, reducing the potential for competition. This suggests that taxonomic floral diversity may be as important as overall floral abundance in enabling the coexistence of bumble bee species. Something about elevation...

As discussed earlier, the interpretation of the patterns in our data hinge strongly on the extent to which interspecific competition actually occurs in our study system. We have argued that the assumption of minimal competition is reasonable in the context of our study, owing to the rich floral resources and harsh abiotic conditions of the Berchtesgadener Alps, together with the manifest co-occurrence of species with similar tongue length. This leads to the interpretation of abundance-conditioned visitation rates as *preferences* and diet overlap as *potential competition*. To the extent that interspecific competition does occur in our system, though, visitation rates would represent floral preference distorted by the pressure of competition toward niche partitioning, rendering diet overlap an underestimate of potential competition. While we maintain, based on our knowledge of our study system, that the first interpretation is more likely, both interpretations are plausible and must be kept in mind.

<!-- We analyze community mean niche overlap as an indication of diffuse competition [@Pianka1974-qk; MacArthur1972-qz] -->

# ACKNOWLEDGEMENTS

>>>>

<!-- Don't forget: I need to fix the Gentiana/Gentianella/Gentianopsis nomenclature problem and rerun all the models before I finalize the paper. This won't change anything about the interpretation of the  analysis, but it needs to be done. It might make Gentiana show up in the PCoA. -->

@@

All analyses were conducted in `R` `r citep(citation())`. Data handling and visualization were performed with the `tidyverse` suite [@Wickham2019-aa]. Bayesian multilevel models were implemented in Stan [@Stan2020aa] via the package `brms` [@Burkner2017-aa] using default priors, and visualizations were prepared with the packages `tidybayes` [@Kay2020-aa] and `ggdist` [@Kay2021-aa]. Ordination analysis and permutation testing were performed with the package `vegan` [@Oksanen_2019] and visualized using the packages `ggplot2` [@Wickham2016-aa] and `ggvegan` [@Simpson2019aa]. Hierarchical generalized additive models (HGAMs) were implemented with the package `mgcv` [@Wood_2017] and visualized with the packages `mgcViz` [@Fasiolo2018aa] and `tidymv` [@Coretta-aa2021]. All R code used in this analysis, together with full model output and diagnostics, is available in Appendix X.

@@

@@

To investigate elevational patterns of *within-site* niche overlap, we modeled community-mean niche overlap as a function of level (morphotype, family, genus) and elevation using HGAMs. Community mean niche overlap, which we interpret as an index of potential *diffuse* competition [@Pianka1974-qk; @MacArthur1972-qz], was calculated by taking the mean niche overlap (within level) across all species pairs for each site. Level was included as both a fixed intercept effect and a smoothing factor in conjunction with elevation, following the type "I" model form described by @Pedersen2019-kf. To account for the proportional nature of our data, we used a beta distribution family with a logit link function.

@@

```{r fig3, echo = FALSE,  fig.cap = paste("Posterior conditional mean visitation probability parsed by Kugler morphotype and bumble bee tongue length class. Point-interval plots depict 66\\% and 95\\% CIs. Letters indicate group differences based on evidence ratio at alpha = 0.05. Note that the range of the x-axis has been rescaled for each panel to emphasize the comparison of visitation probabilities across tongue length classes within floral morphotypes.")}
knitr::include_graphics("../output/brm_pvis00_plot1.png")
```

```{r fig4, echo = FALSE,  fig.cap = paste("Posterior conditional mean visitation probability by Kugler morphotype and bumble bee species. The tongue length class of each species is color-coded. Point-interval plots depict 66\\% and 95\\% CIs. Note that the range of the x-axis has been rescaled for each panel to emphasize the comparison of visitation probabilities across bumble bee species within floral morphptypes.")}
knitr::include_graphics("../output/brm_pvis02_plot1.png")
```

@@

While bumble bees are, indeed, remarkably generalized in their foraging, they are highly discriminating generalists, dramatically over-representing certain floral taxa and morphotypes relative to their abundance while ignoring others. In our study system, lip, flag, and head flowers were by far the most strongly preferred when visitation was pooled across all bumble bee species, while bell, disc, and stalk-disc flowers were underrepresented. Patterns of pooled visitation are, of course, driven disproportionately by the behavior of the most abundant bumble bee species, which in our system were *B. pascuorum*, *B. pratorum*, *B. soroeensis*, *B. terrestris/lucorum*, and *B. wurflenii* (Sponsler et al., submitted). Visitation to lip and flag flowers was driven especially by the medium-tongued *B. pascuorum*, the most abundant bumble bee in our study system, and secondarily by the medium-tongued *B. wurflenii*. Visitation to head flowers was driven mainly by the short-tongued *B. soroeensis*.

@@

As discussed earlier, the interpretation of floral visitation rates and diet overlap hinge strongly on the extent to which interspecific competition actually occurs in our study system. We have argued that the assumption of minimal competition is reasonable in the context of our study, owing to the rich floral resources and harsh abiotic conditions of the Berchtesgadener Alps, together with the manifest co-occurrence of species with similar tongue length. This leads to the interpretation of abundance-conditioned visitation rates as *preferences* and diet overlap as *potential competition*. To the extent that interspecific competition does occur in our system, though, visitation rates would represent floral preference distorted by the pressure of competition toward niche partitioning, rendering diet overlap an underestimate of potential competition. While we maintain, based on our knowledge of our study system, that the first interpretation is more likely, both interpretations are plausible and must be kept in mind.

@@

The question of how bumble bee species with similar tongue lengths coexist hinges on the relationship between species-level floral preferences, trait-matching, and the emergent phenomenon of community-level patterns of diet overlap. While tongue length is a relatively strong predictor of species-level floral preferences, it is at best a weak predictor of diet overlap. This is due in part to divergent preference of bumble bees with similar tongue lengths toward certain floral morphotypes. Perhaps more importantly, though, diet overlap at the level of floral genus tended to be about half as strong as diet overlap at the level of floral morphotype, indicating that bumble bees with similar morphological affinities, driven in part by tongue length compatibility, subdivide floral morphotypes at the genus- and species-level, reducing the potential for competition. This suggests that taxonomic floral diversity may be as important as overall floral abundance in enabling the coexistence of bumble bee species. **Something about elevation... Somemthing that broadens back out to the general themes of species coexisitence and climate change...**

<!-- Maybe add something somewhere about MacArthur's insight regarding the importance of transient and sporadic competition as a driver community assembly. -->

# ACKNOWLEDGMENTS
